Introducing an Ecological Trait-data Standard

### 2017-11-08 manuscript/manuscript.Rmd
fixing bibliography

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions [@Forsman2015] (@refs).

@@

Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13; @diaz16]. Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13] which may be used to inform ecological networks [@munoz17; @weiner14; @miranda13; @digel11] and parameterize ecological population-dynamic models [@schneider12].   
On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17]. Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14; @gravel17; @lajoje15] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].  Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].  
In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.

@@

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/). Besides these databases, an unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> @Sokol2011, @Price2013 <!-- Nadja: oldest relevant pub. in Dryad-->]. In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g. @ricklefs17; @amphibio17; @iversen17; @myhrvold15; @kissling14; @jones09]. Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is lacking. It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@refs].

@@

While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored. 
The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr15], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczoreck12), relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr15]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

@@

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (a.k.a. wide-table format,  Fig. 2*a*), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as 'NA'. It may store additional information (e.g. on variation of means or literature source) in secondary colums or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10]. Also at this level of information, missing entries, e.g. about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15]. However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnik15].

@@

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr15], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence). This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.  Longtable dataset purport multiple advantages for the data manipulation (e.g. filtering, subsetting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modeling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. They further draw from a defined set of columns, which simplifies merging of datasets.

@@

Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements. The Handbook for Measurement of traits [@moretti17] is a model example of such a methodological standardization approach.

@@

For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt'). EOL TraitBank proposes the use of Darwin Core Archives [http://eol.org/info/structured_data_archives, DwC-A; @robertson09] to upload such related data along with the core trait dataset, which is also the accepted format for GBIF [@braak11]. These are .zip archives containing data table txt-files along with a descriptive metadata file (in .xml format). Detailled descriptions and tools for validation can be found on the website of EOL (http://eol.org/info/cp_archives) and GBIF (http://tools.gbif.org/dwca-assistant/).

@@

Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15a; @ricklefs17; @vanderplas17]. 
To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB. The package  'TR8' provides similar access to plant traits from a list of databases (including LEDA, BiolFlor and Ellenberg Values)[@bocci15] and aggregates them into a species-trait matrix. However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.  
We provide a script (see online appendix and R-package vignette) to demonstrate the harmonization of these different data sources into our data standard.

>>>>

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07] and have been used to describe the role of species in ecosystems [@fontaine05] or their ability to persist under certain environmental conditions [@forsman15].

@@

Traits have proven useful for bridging individual level physiology and behaviour to processes at the ecosystem scale and patterns of biodiversity at the global level [@diaz13; @diaz16]. Comparative study of traits informs researchers about physiological and evolutionary trade-offs, and helps identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. For instance, defense traits and predator feeding traits have co-evolved, just as pollinator and flower traits. This 'trait-matching' enables the inferrence of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13] which may be used to inform ecological networks [@munoz17; @weiner14; @miranda13; @digel11] and parameterize ecological population-dynamic models [@schneider12].   
On the level of species assemblages, community weighted means or variance of functional traits [@laliberte10a] or multivariate measures of 'functional diversity' [@petchey09; @villeger08a] are used to understand changes in functional composition or the loss of functional diversity in consequence to global change or local anthropogenic land use [environmental filtering; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17]. Since traits are closely related to ecological mechanisms, such patterns in functional community composition are promising versatile statistical tools to assess the current state of an ecosystem [ @moretti13; @kuhsel15;  @gamez-virues15; @birkhofer14; @gravel16; @lajoie15] or of ecosystem services [@wood15b; @bello10; @diaz07; @lavorel12; @allan15].  Due to the abovementioned trait-matching, these analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16].  
In recent years, functional ecology has thus grown into a major field of ecological research, linking basic physiological knowledge to functions and services as well as applied matters in land use management and biodiversity conservation, as well as risk assessment of anthropogenic stressors and climate change.

@@

In the animal kingdom, specialised trait-databases have been created that cover different organism groups, interaction types, or ecosystems. For instance, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/). Besides these databases, an unknown plethora of single datasets has been published along with article publications [e.g., @gossner15, @gossner15c, @renner14; @birkhofer17; @vanderplas17  <!-- FDS: add some classics of trait data --> @sokol11, @price13 <!-- Nadja: oldest relevant pub. in Dryad-->]. In the past years, many datasets have been aggregated and harmonized into greater collections with particular organismal focus [e.g. @ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09]. Unlike for plants, a single unified platform and harmonizing scheme for animal trait data is lacking. It has been pointed out that data on animals, covering multiple trophic levels and having evolved very different body plans will not be a simple task and go far beyond the efforts produced for plant traits [@refs].

@@

While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored. 
The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczorek12), relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

@@

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (a.k.a. wide-table format,  Fig. 2*a*), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as 'NA'. It may store additional information (e.g. on variation of means or literature source) in secondary colums or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15]. However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnick16].

@@

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence). This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.  Longtable dataset purport multiple advantages for the data manipulation (e.g. filtering, subsetting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modeling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. They further draw from a defined set of columns, which simplifies merging of datasets.

@@

Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements. The Handbook for Measurement of traits [@moretti17a] is a model example of such a methodological standardization approach.

@@

For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt'). EOL TraitBank proposes the use of Darwin Core Archives [http://eol.org/info/structured_data_archives, DwC-A; @robertson09] to upload such related data along with the core trait dataset, which is also the accepted format for GBIF [@gbif17]. These are .zip archives containing data table txt-files along with a descriptive metadata file (in .xml format). Detailled descriptions and tools for validation can be found on the website of EOL (http://eol.org/info/cp_archives) and GBIF (http://tools.gbif.org/dwca-assistant/).

@@

Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15c; @ricklefs17; @vanderplas17]. 
To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB. The package  'TR8' provides similar access to plant traits from a list of databases (including LEDA, BiolFlor and Ellenberg Values)[@bocci15] and aggregates them into a species-trait matrix. However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.  
We provide a script (see online appendix and R-package vignette) to demonstrate the harmonization of these different data sources into our data standard.
